Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-12-06 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
update pdf
some code changes to be able to run

```{r patterns, fig.height=1.5, fig.width=4.1, fig.cap = "Missingness in multilevel data", echo=FALSE}
dat <- expand.grid(rows = 1:7, cols = 1:6) %>% 
  cbind(text = c("1", "1", "2", "2", "3", "", "N", rep("", 35)),
        miss = c(rep("", 16), "NA", "NA", "", "", "", "NA", "", "", "NA", rep("", 17)))
ggplot(dat, aes(x = cols, y = rows)) +
  geom_tile(fill = "white", color = "black", size = 0.5) +
  geom_text(aes(label = text), color = "black", size = 3) +
  geom_text(aes(label = miss), color = mice:::mdc(2), family = "mono", fontface = "bold") + 
  scale_x_continuous(
    breaks = 1:6,
    labels = c("cluster", expression(X[1]), expression(X[2]), expression(X[3]), "...", expression(X[p])),
    name = NULL,
    position = "top"
  ) +
  scale_y_continuous(breaks = 1:7, labels = c(1:5, "...", "n"), name = NULL, trans = "reverse") +
  coord_cartesian(expand = c(0,0,0,0)) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        plot.margin = unit(c(0,0,0,0), "pt"))
# TODO: use math font for n and N
```
<!-- Column $X_1$ in Figure 1 is completely observed, column $X_2$ is systematically missing in cluster 2, and column $X_3$ is sporadically missing. To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account. For example, the sporadic missingness in $X_3$ could be easily amended if this would be a cluster-level variable (and thus constant within clusters). We could then just extrapolate the true (but missing) value of $X_3$ for unit 1 from unit 2, and the value for unit 4 from unit 3. If $X_3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but would need to use some kind of missing data method, or discard the incomplete units altogether (i.e., complete case analysis). Complete case analysis can however introduce bias in statistical inferences and lowers statistical power. Further, with the systematic missingness in $X_2$, it would be impossible to fit a multilevel model without accommodating the missingness in some way. Complete case analysis in that case would mean excluding the entire cluster from the analyses. The wrong choice of missing data handling method can thus be extremely harmful to the inferences.  -->

@@

Impute the data:
```{r}
imp <- mice( hivdata, # dataset with missing values
                        m = 10,   # number of imputations
                        maxit = 1,
                        seed = 1234, #seed attached to the dataID
                        meth = meth, #imputation method vector
                        pred = pred, #imputation predictors matrix
                        print = T,
                        meta_method="reml",
                        pmm=FALSE)
```

>>>>

```{r patterns, fig.height=1.5, fig.width=4.1, fig.cap = "Missingness in multilevel data", echo=FALSE}
dat <- expand.grid(rows = 1:7, cols = 1:6) %>% 
  cbind(text = c("1", "1", "2", "2", "3", "", "N", rep("", 35)),
        miss = c(rep("", 16), "NA", "NA", "", "", "", "NA", "", "", "NA", rep("", 17)))

ggplot(dat, aes(x = cols, y = rows)) +
  geom_tile(fill = "white", color = "black", size = 0.5) +
  geom_text(aes(label = text), color = "black", size = 3) +
  geom_text(aes(label = miss), color = mice:::mdc(2), family = "mono", fontface = "bold") + 
  scale_x_continuous(
    breaks = 1:6,
    labels = c("cluster", expression(X[1]), expression(X[2]), expression(X[3]), "...", expression(X[p])),
    name = NULL,
    position = "top"
  ) +
  scale_y_continuous(breaks = 1:7, labels = c(1:5, "...", "n"), name = NULL, trans = "reverse") +
  coord_cartesian(expand = c(0)) + #Changed to run GV
  # coord_cartesian(expand = c(0,0,0,0)) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        plot.margin = unit(c(0,0,0,0), "pt"))
# TODO: use math font for n and N
```
<!-- Column $X_1$ in Figure 1 is completely observed, column $X_2$ is systematically missing in cluster 2, and column $X_3$ is sporadically missing. To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account. For example, the sporadic missingness in $X_3$ could be easily amended if this would be a cluster-level variable (and thus constant within clusters). We could then just extrapolate the true (but missing) value of $X_3$ for unit 1 from unit 2, and the value for unit 4 from unit 3. If $X_3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but would need to use some kind of missing data method, or discard the incomplete units altogether (i.e., complete case analysis). Complete case analysis can however introduce bias in statistical inferences and lowers statistical power. Further, with the systematic missingness in $X_2$, it would be impossible to fit a multilevel model without accommodating the missingness in some way. Complete case analysis in that case would mean excluding the entire cluster from the analyses. The wrong choice of missing data handling method can thus be extremely harmful to the inferences.  -->

@@

Impute the data:
```{r eval = FALSE}
imp <- mice( hivdata, # dataset with missing values
                        m = 10,   # number of imputations
                        maxit = 1,
                        seed = 1234, #seed attached to the dataID
                        meth = meth, #imputation method vector
                        pred = pred, #imputation predictors matrix
                        print = T,
                        meta_method="reml",
                        pmm=FALSE)
```
